Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-16 manuscript/manuscript.tex
revisions to abstract and Results section 1

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

@@

\begin{abstract}
\noindent  
Molecular phylogenetics is used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that substitution models informed by experimental measurements of the purifying selection \skhcomment{due to site-specific amino-acid preferences} lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites \skhcomment{repeat sentence start}.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times. 
\end{abstract}

@@

\subsection*{Different ways hat substitution models account for purifying selection}

\jdbcomment{Some other comments on this section, which I think in general is pretty good: $\omega$ shows up in the figure, but is never mentioned in the text. I feel like GY94 models need to be explained at least in terms of what they and $\Gamma$ distribution stand for. Just introducing nomenclature. I feel like the last paragraph should tie back to panel C of the figure.}

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling the site-to-site rate variation of purifying selection.}
(A) The implied dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary following some statistical distribution (blue circles). 
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here.
(B) A substitution model stationary state defines the expected sequence composition after a very long evolutionary time. 
Depending on how the model is parametrized, this expected sequence composition cab be uniform or variable across sites. 
Each column is a site in the protein and the height of each letter is the expected proportion of that amino acid at stationary state. 
(C) Substitution models can both, one, or neither of these features and we use models from the GY94 and ExpCM families to represent all possible combinations. 
Green indicates presence of a feature and white indicates absence. 
}
\end{figure}

@@

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
For example, in GY94+$\Gamma\omega$ (M5 of \cite{yang2000codon}), the implied dN/dS parameter, $\omega$, follows a gamma distribution (\ref{fig:model_feature}A).
Under GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
However, the specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting \skhcomment{cite rodrigue paper}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Finally, these two strategies to account for purifying selection, $\Gamma\omega$ rate variation and site-specific stationary states, are not mutually exclusive. 
Some proteins may be better modeled as a combination of the two. 
Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to dissect the effect of each model of purifying selection on branch length estimation.

>>>>

\topmargin 0.0cm
%\oddsidemargin 2cm
%\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

@@

\begin{abstract}
\noindent  
Molecular phylogenetics is used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times. 
\end{abstract}

@@

\subsection*{Different ways that substitution models account for purifying selection}

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling the site-to-site rate variation of purifying selection.}
(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary following some statistical distribution (blue circles). 
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin. 
A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
(B) A substitution model stationary state defines the expected sequence composition after a very long evolutionary time. 
Most substitution models have stationary states that are uniform across sites.
However, substitution models can have site-specific stationary states.
In the logo plots, each column is a site in the protein and the height of each letter is the frequency of that amino acid at stationary state. 
(C) Substitution models can incorporate neither, one, or both of these features.
Here we will use substitution models from the \citet{goldman1994codon} (GY94) and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed rates to represent all possible combinations. 
%Green indicates presence of a feature and white indicates absence. 
}
\end{figure}

@@

Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use\jdbcomment{find a citation for this and replace word ``accurate'' with something else if appropriate. Perhaps this one or something cited within \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4620419/}}.
Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}, which we denote GY94.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called GY94 M0.
This gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

A single gene-wide $\omega$ ignores the fact that purifying selection is heterogeneous across sites.
The most common strategy to ameliorate this defect is to allow the rate of nonsynonymous change ($\omega$) to vary among sites following some statistical distribution~\citep{yang1994maximum,yang2000codon}.
For instance, in the M5 variant of the GY94 model~\citep{yang2000codon}, $\omega$ follows a gamma distribution as shown in \ref{fig:model_feature}A.
We will denote this model as GY94+$\Gamma\omega$.
A GY94+$\Gamma\omega$ captures the fact that the rate of nonsynonymous change can vary across sites. 
However, this formulation treats all nonsynonymous substitutions equivalently, since the rate is agnostic to the identity of the specific amino-acid mutation.

A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary}.
These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94 M5 model with the typical F3X4 stationary state\jdbcomment{add citation, maybe just GY94 paper?} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data\skhcomment{cite rodrigue paper}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter can be drawn from a distribution just like as for GY94-style models. 
We will denote ExpCM's that also include a distribution over $\omega$ as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.
